---
title: "Survival analysis"
author: "Tomoya Ozawa"
date: "2023-07-07"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Non parametric model

- we do not assume any distribution assumption and do not use covariates.

## Set up

[Reference](https://note.com/maxwell/n/nc78c55afe944?magazine_key=mc84c90fb0e35)

```{r, message=FALSE}
library(tidyverse)
library(MASS)
library(survival)
library(survminer)
```



```{r}
data(gehan)
gehan
```

- this data set is identified with pair and treat
- columns:
  - `time`: duration individual exists
  - `cens`: takes 1 if the observation is censored

```{r}
cut.points <- gehan %>% 
  filter(cens == 1) %>% 
  arrange(time) %>% 
  pull(time) %>% 
  unique()

cut.points
```



```{r}
gehan2 <- survival::survSplit(data = gehan, 
                              cut = cut.points,
                              start = "start",
                              end = "time",
                              event = "cens")
colnames(gehan2)[4] <- "stop"
gehan2
```


## Kaplan-Meier Curve

- compute the survival function, $S(t) = 1-F(t)$

```{r}
# create object for survival analysis
surv.obj <- Surv(time = gehan$time, event = gehan$cens)

# estimate Kaplan-Meier curve for each value in specified variable
# in this case, plot the curve for treat or not
ge.sf <- survfit(surv.obj ~ treat, data = gehan)

# dataframe showing the estimation result
ge.sf.df <- survminer::surv_summary(ge.sf, data = gehan)
ge.sf.df
```


- columns:
  - `n.risk`: the number of existing people
  - `n.event`: the number of death
  - `n.censor`: the number of people who were subjected to the censoring
  
$$
S(t_{f-1}) = \prod_{i = 1}^{f-1}Pr(T>t_i | T \geq t_i) \ , \quad
S(t_f) = S(t_{f-1})Pr(T>t_f | T \geq t_f)
$$

For treated group,

$$
S(t_1) = \frac{21-3}{21} = 0.857\ldots \quad S(t_2) = S(t_1)\times  \frac{17-1}{17} = 0.806\ldots
$$

Here we assume independence of censoring, which means we regard censored observations have the same pattern with the other existing observations.

```{r}
survminer::ggsurvplot(
   fit = ge.sf,
   data = gehan,
   conf.int = T,
   pval = T,
   risk.table = T,
   cumevents = T,
   cumcensor = T,
   ggtheme = ggplot2::theme_light(),
   tables.height = 0.15
)
```


## Log-rank test

$$
H_0: \text{ There is no difference between two KM curves}
$$

```{r}
survdiff(surv.obj ~ treat, data = gehan)
```

Since p-value is less than 0.01, we reject the null hypothesis with the 1% level of significance.



# Semi-nonparametric model

- We do not assume any distribution assumption, but we employ covariates.

## Cox PH model

- Cox PH model is a way to model a hazard function.
- PH stands for Proportional Hazard
- Hazard function indicates the probability of occurring an event given the individual has survived at time $t$.
- In Cox PH model, the hazard function depends on covariates, $X$, as well as time, $t$.

$$
h(t, X) = h_0(t) \ \text{exp}\left[\sum_{i = 1}^{p}\beta_iX_i\right]
$$


- One of the important feature of this model is that the proportion of the values of the hazard function in two different covariates is constant and independent of time. **Proportional hazard Assumption**

```{r}
# add logWBC as a covariate
gehan <- read_csv("https://raw.githubusercontent.com/Greenwind1/survival-analysis/master/input/gehan-wbc.csv",
                  col_types = cols(treat = col_character())) %>% 
  mutate(treat = ifelse(treat == "6-MP", "6MP", "control"))
```

```{r}
surv.obj <- Surv(time = gehan$time, event = gehan$cens)
m1 <- coxph(surv.obj ~ treat + logWBC, 
           data = gehan,
           method = "efron")
```

- plot of adjusted survival curve usiusinggn the mean value of covariate

```{r}
survminer::ggsurvplot(
 fit = survfit(m1, conf.int = 0.95),
 data = gehan,
 conf.int = T,
 ggtheme = ggplot2::theme_light(),
)
```

- Plot showing the difference caused from treatment

```{r}
new.data <- data.frame(
 treat = c("control", "6MP"),
 logWBC = rep(mean(gehan$logWBC), 2)
)

survminer::ggsurvplot(
 fit = survfit(m1,newdata = new.data, conf.int = 0.95),
 data = gehan,
 conf.int = T,
 legend.labs = c("control", "6MP"),
 ggtheme = ggplot2::theme_light(),
)
```

- we can see the survival curve for each value of covariate by specifying their values as new data.


- To validate the Proportional hazard assumption, we would do

1. 














